Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-04 draft.tex
shortened 'data set' section

\noindent Our approach allows the simultaneous analysis of many variables.

@@

Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.

@@

The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.

@@

The data set comprises 134 variables (and 9 additional image-derived variables described in the next subsection):

@@

	\item 21 metric variables, mostly describing somatometric variables and markers retrieved from blood analysis

@@

Many variables comprise missing data.
Some variables are follow-up questions, covering the treatment of a disease.
Others are exclusive to subgroups, such as women-specific variables, e.g. \emph{menstrual status}.
Sparse variables are statistically less resilient and need to be treated with care during the analysis.

@@

	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.

@@

	(a) right: FEM aligned to an MRI data set.

@@

	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.

@@

	(b) Using the Frenet frame, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).

@@

	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.

@@

	We also extracted the variables for each projection axis to assess their information gain.

@@

Four trained radiologists acquired the medical image data for each subject in a standardized way using a 1.5~Tesla Magnetic Resonance Imaging (MRI) scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).

@@

The spine protocol consisted of a sagittal $T_1$- and $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.

@@

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-defined landmarks.
%%
A click on the L3 vertebra center initializes the model, two clicks on the top and bottom of the vertebra describe an initial height estimation (Fig.~\ref{fig:centerline}~a).
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
%%
Each subject's model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
rotation and scaling.
%%
The detection fails for several subjects due to imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ out of $6,753$ lumbar spine models were obtained and constitute the foundation of this work.
%%

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).

@@

Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:

@@

These metrics are also extracted in the sagittal, coronal and transversal projection of the model.
We assess the information gain of each dimension using these projections.
This yields a total of 9 image-derived variables.

>>>>

% \noindent Our approach allows the simultaneous analysis of many variables.

@@

%Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects

@@

%The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.

@@

The data set comprises 134 variables: %(and 9 additional image-derived variables described in the next subsection):

@@

	\item 21 metric variables, describing somatometric variables and markers retrieved from blood analysis

@@

% Many variables comprise missing data.
% Some variables are follow-up questions, covering the treatment of a disease.
% Others are exclusive to subgroups, such as women-specific variables, e.g. \emph{menstrual status}.
% Sparse variables are statistically less resilient and need to be treated with care during the analysis.

@@

	(a) Finite element model (FEM) of the lumbar spine (left), capturing the L1-L5 vertebrae and the lumbar spine canal (right).

@@

%	(a) right: FEM aligned to an MRI data set.

@@

	%The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
	The purple dashed line describes the lumbar spine canal centerline with 92 points.

@@

	(b) We extract the weighted sum of \emph{curvature} and \emph{torsion} for all 92 points (white dashes).

@@

	The \emph{curvature angle} ($\alpha$) is calculated using the top, bottom and middle point.

@@

	We extracted the variables for each projection axis to assess their information gain.

@@

Magnetic Resonance Imaging (MRI) scans were obtained for each subject \cite{Hegenscheid2013}.

@@

% Four trained radiologists acquired the medical image data for each subject in a standardized way using a 1.5~Tesla Magnetic Resonance Imaging (MRI) scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
% The spine protocol consisted of a sagittal $T_1$- and $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.

@@

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.

@@

%The tetrahedron-based finite element model (Fig.~\ref{fig:centerline}~a) is initialized using three user-defined landmarks.
%A click on the L3 vertebra center initializes the model, two clicks on the top and bottom of the vertebra describe an initial height estimation (Fig.~\ref{fig:centerline}~a).
The tetrahedron-based finite element model (Fig.~\ref{fig:centerline}~a) captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
%The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
%Each subject's model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, rotation and scaling.
The detection fails for several subjects due imaging artifacts or strongly deformed spines, yielding models for $2,540$ out of $6,753$ subjects.
%$2,540$ out of $6,753$ lumbar spine models were obtained and constitute the foundation of this work.
%We have to assess the model accuracy to extract key figures from it.
The detection model depicts the vertebrae positions, spine canal curvature, but lacks detailed information about their volume.
%It captures reliable information about spine canal curvature.
In \cite{Klemm2013VMV}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b): %\cite{Frenet}:

@@

% These metrics are also extracted in the sagittal, coronal and transversal projection of the model.
% We assess the information gain of each dimension using these projections.
% This yields a total of 9 image-derived variables.
These metrics are also extracted in the sagittal, coronal and transversal projection of the model, yielding 9 image-derived variables.
